---
title: "Homework 1: Using OLS Regression to Predict Median House Values in Philadelphia"
authors: Chloe Alimurong, Sophia Meyers & Evelyn Zhong
date: October 5, 2026
output: rmdformats::readthedown
---

# Introduction
In this analysis, we will assess the relationship between median house values in Philadelphia and variables such as # Households Living in Poverty, % with Bachelor's Degree or Higher, % Vacant Houses and % Single House Units. 

The predictors in our analysis describe different aspects of housing that could potentially influence house values and prices. The number of households in poverty and with a bachelors degree, to some degree, explain the economic status of the demographic's housing market. Those living in poverty may not be able to afford houses with higher price tags and those with a bachelors degree may have an income that could afford houses with higher price tags. These population factors indirectly relate to house value, thus why we chose them to predict in our analysis. For unit-specific predictors like vacant houses and single house units, these two signal environmental factors that could influence house value. For vacant houses, we expect that the neighborhood may not be as desirable or may not have high economic activity, making it less desirable to live, which could decrease house values. The percentage of single house units may also highlight the neighborhood's population. In Philadelphia, it is rare to see single-family detached homes. The common house type are multi-family style houses. So for those living in single-family homes, it indicates the desire to live at longer terms and farther from the city center. These types of homes may single out those with higher incomes to afford a single unit house within a city, which may mean higher house value. All of these predictors indicate something about the housing market population and environment indirectly, making them good examples of variables to predict median house value within Philadelphia.

```{r}
#| label: setup
#| warning: false
#| message: false 
# note to self: make sure all libraries are in use or necessary (since tidyverse has a lot)...

library(sf)
library(Hmisc)
library(DAAG)
library(car)
library(MASS)
library(rsq)
library(tidyverse)
library(caret)
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
library(corrplot)

```

# Exploratory Analysis
## Data Cleaning
The original Philadelphia Block Group dataset had 1816 observations and was cleaned in order to achieve a dataset with 1720 observations. The following items were removed; Block groups where population < 40, Block groups where there are no housing units, Block groups where the median house value is lower than 10,000, One North Philadelphia block group which had a very high median house value (over 800,000), and a very low median household income (less than $8,000). The purpose of their removal was to remove any outliers, no data, and insufficient sample size from our analysis.


## Methods
In this section, we will evaluate the summary statistics and distributions of the dependent variable and predictors.
In our exploratory data analysis, we will also examine the correlations between the predictors. A correlation is the relationship between two variables. The sample correlation coefficient r, the strength of a linear relationship between two variables (correlation strength) on a -1 to 1 scale, can be calculated with the following equation:
$$r = \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum (x_i - \bar{x})^2 \sum (y_i - \bar{y})^2}}$$
The possible range of r values is -1 to 1. A correlation of 0 means there is no linear relationship between the two variables.
```{r}
#| label: summary statistics & examining the distribution of the dependent variable and predictors

#regdata <- read.csv("~/GitHub/MUSA-5000-portfolio/HW1/RegressionData.csv")
regdata <- read.csv("RegressionData.csv")
#names(regdata) # gets the names of each variable.

regdata <- regdata %>% mutate(across(where(is.character), as.numeric)) # some of them were returning a string, we need a number to run summary statistics.

vars <- c("MEDHVAL","PCTBACHMOR", "NBELPOV100", "PCTVACANT", "PCTSINGLES")

sum_table <- data.frame(
  Variable = c("Median House Value",
                        "% with Bachelor's Degree or Higher",
                        "# Households Living in Poverty",
                        "% Vacant Houses",
                        "% Single House Units"),
  Mean = sapply(regdata[vars], mean, na.rm = TRUE), #creates column of mean values for each variable, ignoring null values.
  SD = sapply(regdata[vars], sd, na.rm = TRUE) #creates column of sd values for each variable, ignoring null values.
)

sum_table

#running hist on all of the variables we defined above
for (v in vars){
  hist(regdata[[v]],
       main = paste("Distribution of ", v),
       xlab =v,
       col = "lightblue",
       border = "darkblue")
}

```

## Results
According to the summary statistics table, the average Median House Value in Philadelphia, the dependent variable, across the blocks in our sample is 66287.73 with a standard deviation of 60006.08. For the predictor variables: The average % of households with a Bachelors degree or higher is 16.08% with a standard deviation of 17.77%. The average number of households living in poverty is 189 with a standard deviation of 164. The average % Vacant Houses is 11.29% with a standard deviation of 9.63%. The average % Single House Units is 9.23% with a standard deviation of 13.25%.	

```{r}
#| label: log transformations and its effect on the distribution

regdata <- regdata %>%
  mutate(LNMEDHVAL    = log(MEDHVAL),
         LNPCTBACHMOR = log(1 + PCTBACHMOR),
         LNBELPOV100 = log(1 + NBELPOV100),
         LNPCTVACANT  = log(1 + PCTVACANT),
         LNPCTSINGLES = log(1 + PCTSINGLES))

hist(regdata$MEDHVAL, main = "MEDHVAL", xlab = "MEDHVAL",col = "lightblue",border = "darkblue")
hist(regdata$LNMEDHVAL, main = "LNMEDHVAL", xlab = "LNMEDHVAL",col = "lightblue",border = "darkblue")


hist(regdata$PCTBACHMOR, main = "PCTBACHMOR",  xlab = "PCTBACHMOR",col = "lightblue",border = "darkblue")
hist(regdata$LNPCTBACHMOR, main = "LNPCTBACHMOR", xlab = "LNPCTBACHMOR",col = "lightblue",border = "darkblue")


hist(regdata$NBELPOV100,  main = "NBELPOV100",  xlab = "NBELPOV100",col = "lightblue",border = "darkblue")
hist(regdata$LNBELPOV100, main = "LNBELPOV100", xlab = "LNBELPOV100",col = "lightblue",border = "darkblue")

hist(regdata$PCTVACANT,  main = "PCTVACANT",  xlab = "PCTVACANT",col = "lightblue",border = "darkblue")
hist(regdata$LNPCTVACANT, main = "LNPCTVACANT", xlab = "LNPCTVACANT",col = "lightblue",border = "darkblue")

hist(regdata$PCTSINGLES,  main = "PCTSINGLES",  xlab = "PCTSINGLES",col = "lightblue",border = "darkblue")
hist(regdata$LNPCTSINGLES, main = "LNPCTSINGLES", xlab = "LNPCTSINGLES",col = "lightblue",border = "darkblue")

```
The variables are were not normal before the logarithmic transformation, they were mostly skewed right. Whereas after the logarithmic transformation, they become more normally distributed.
In our regression, we use the log-transformed variables for the dependent variable (median house value) and the number of households below poverty predictor.
The other regression assumptions will be examined in a separate section below (Regression Assumption Checks).

```{r}
#| label: Scatter plot for each predictor
#| message: false
#| warning: false

regdata %>%
  dplyr::select(LNMEDHVAL, PCTBACHMOR, LNBELPOV100, PCTVACANT, PCTSINGLES) %>%
  pivot_longer(-LNMEDHVAL, names_to = "Predictor", values_to = "Value") %>%
  ggplot(aes(x = Value, y = LNMEDHVAL)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "purple") +
  facet_wrap(~Predictor, scales = "free_x") +
  labs(x = NULL, y = "LNMEDHVAL")

```

```{r}
#| label: Pearson correlation for each predictor

#cor(regdata$PCTVACANT, regdata$MEDHVAL, method = "pearson")
#cor(regdata$PCTSINGLES, regdata$MEDHVAL, method = "pearson")
#cor(regdata$PCTBACHMOR, regdata$MEDHVAL, method = "pearson")
#cor(regdata$NBELPOV100, regdata$MEDHVAL, method = "pearson")

var <- c("MEDHVAL", "PCTBACHMOR", "NBELPOV100", "PCTVACANT", "PCTSINGLES")

pearson <- rcorr(as.matrix(regdata[, var]), type = "pearson")          
pearson$r
corrplot(
  pearson$r, type = "upper", order = "hclust", diag = FALSE,
  p.mat = pearson$P, sig.level = 0.05, insig = "blank",
  tl.col = "black", tl.srt = 45,
  col = colorRampPalette(c("red", "white", "blue"))(200)
)
```
The correlation matrix shows that there is severe multicollinearity. Especially between Median House Value and Percent of households with bachelors degree or more, matching the conclusion from our predictor maps above.

```{r}
#| label: Chloropleth maps of each variable using the sf (simple features) and ggplot packages
#| note: readOGR has been officially retired as a part of the rgdal library.

#shapefile <- "~/GitHub/hw-1-musa5000/Lecture 1 - RegressionData.shp/RegressionData.shp"
shapefile <- "Lecture 1 - RegressionData.shp/RegressionData.shp"

c_map = read_sf(shapefile)

#names(c_map)

c_map <- c_map %>%
  mutate(LNMEDHVAL    = log(MEDHVAL),
         LNBELPOV100 = log(1 + NBelPov100),
         PCTVACANT, PCTSINGLES, PCTBACHMOR)

#maps the log of the dependent variable: LNMEDHVAL
ggplot(c_map) +
  geom_sf(aes(fill=LNMEDHVAL))+
  scale_fill_distiller(palette = "BuPu", direction = 1)+
  theme_void()+
  labs(title = "Log of Median House Value")

# maps 2-5: the four predictors
p1 <- ggplot(c_map) + geom_sf(aes(fill = LNBELPOV100)) +scale_fill_distiller(palette = "BuPu", direction = 1) + theme_void() + labs(title = "LNBELPOV100")
p2 <- ggplot(c_map) + geom_sf(aes(fill = PCTBACHMOR))   +scale_fill_distiller(palette = "BuPu", direction = 1) + theme_void() + labs(title = "PCTBACHMOR")
p3 <- ggplot(c_map) + geom_sf(aes(fill = PCTVACANT))    +scale_fill_distiller(palette = "BuPu", direction = 1) + theme_void() + labs(title = "PCTVACANT")
p4 <- ggplot(c_map) + geom_sf(aes(fill = PCTSINGLES))   +scale_fill_distiller(palette = "BuPu", direction = 1) + theme_void() +labs(title = "PCTSINGLES")

# combine into one 2x2 figure
(p1 | p2) / (p3 | p4)
```
The only maps that look similar are the dependent variable map (log of the median house value) and the percent of households with a bachelors degree or more. The rest of the maps look vastly different. The predictor we expect to be strongly associated with the dependent variable based solely on visualization is the percent of households with a bachelor's degree and more. The striking similarity of these two maps signals to us that there may be inter-correlation between these variables. We also suspect multicollinearity to be an underlying issue for this similarity.

##Sophia
# Regression Analysis
## Methods
### Multiple Regression Analysis
Multiple linear regression models how a continuous dependent variable changes with several predictors at once. Our model is:

$$LNMEDHVAL = \beta_0 + \beta_1 PCTVACANT + \beta_2 PCTSINGLES + \beta_3 PCTBACHMOR + \beta_4 LNNBELPOV100 + \varepsilon$$

where $\beta_0$ is the intercept, each $\beta_i$ is the expected change in LNMEDHVAL for a one-unit change in that predictor holding the others constant, and $\varepsilon$ is the random error. LNMEDHVAL and LNNBELPOV100 are natural-log transformed. The model assumes linearity, independent observations, normally distributed residuals, constant error variance (homoscedasticity), and no severe multicollinearity among the predictors.

The parameters to estimate are $\beta_0, \dots, \beta_4$ and the error variance $\sigma^2$, the variance of $\varepsilon$. The coefficients are estimated by least squares, which minimizes the sum of squared residuals: $\hat{\boldsymbol{\beta}} = (X^{T}X)^{-1}X^{T}y$.

The coefficient of multiple determination is $R^2 = SSR/SST = 1 - SSE/SST$, where SSE is the error sum of squares, SSR the regression sum of squares, and SST the total sum of squares. The adjusted $R^2 = 1 - \frac{SSE/(n-k-1)}{SST/(n-1)}$ penalizes the number of predictors $k$ ($n$ is the number of observations).

We test the overall model with the F-ratio, $F = \frac{SSR/k}{SSE/(n-k-1)}$, where $H_0: \beta_1 = \dots = \beta_4 = 0$ and $H_a$: at least one $\beta_i \neq 0$. For each predictor, we test $H_0: \beta_i = 0$ against $H_a: \beta_i \neq 0$ using a t-test.

### Additional Analyses
Stepwise regression adds or removes predictors one at a time, keeping changes that lower the Akaike Information Criterion (AIC = $2p - 2\ln(\hat{L})$, where $p$ is the number of parameters and $\hat{L}$ is the maximized likelihood). Its limitations are that it is greedy and may miss the best model, it can overfit, its p-values are overly optimistic, and it ignores theory.

K-fold cross-validation (k = 5) tests how well a model predicts data it was not fit to. The data are split into five folds. Each fold is held out in turn while the model is fit to the other four, and the model then predicts the held-out fold. Models are compared using the root mean square error, $RMSE = \sqrt{\frac{1}{n}\sum (y_i - \hat{y}_i)^2}$, where $\hat{y}_i$ is the cross-validated prediction. A smaller RMSE means better prediction.

### Software
All analyses were performed in R.

## Results
```{r}
#| label: Using Stepwise regression to determine the best model based on the Akaike Information Criterion

model_full <- lm(
  LNMEDHVAL ~ PCTVACANT + PCTSINGLES + PCTBACHMOR + LNBELPOV100,
  data = regdata
)

summary(model_full)

anova(model_full)
```
### Regression Results
All four predictors are statistically significant (p < 0.0001). Because the dependent variable is logged, coefficients are approximate percentage changes in median house value, holding the other predictors constant:

PCTVACANT ($\hat{\beta}_1$ = -0.0192): each additional percentage point of vacant housing is associated with about a 1.90% decrease in median house value.
PCTSINGLES ($\hat{\beta}_2$ = 0.0030): each additional percentage point of single-family homes is associated with about a 0.30% increase.
PCTBACHMOR ($\hat{\beta}_3$ = 0.0209): each additional percentage point of residents with a bachelor's degree or higher is associated with about a 2.11% increase.
LNBELPOV100 ($\hat{\beta}_4$ = -0.0789): a 1% increase in the number of households in poverty is associated with about a 0.078% decrease.

For each predictor, the p-value is the probability of an estimate this large if the true coefficient were zero, so we reject $H_0: \beta_i = 0$ for all four. The model explains about 66% of the variance in LNMEDHVAL ($R^2$ = 0.6623, adjusted $R^2$ = 0.6615). The F-ratio is 840.9 on 4 and 1715 degrees of freedom (p < 0.0001), so we reject the null that all coefficients are zero.

### Regression Assumption Checks
We examined the variable distributions earlier. The scatter plots of LNMEDHVAL against each predictor do not look cleanly linear, since PCTVACANT, PCTSINGLES and PCTBACHMOR all have clusters at or near zero and PCTSINGLES is only weakly related to LNMEDHVAL. Standardized residuals are residuals divided by their estimated standard deviation, so values beyond $\pm 2$ are unusual and beyond $\pm 3$ are potential outliers. The histogram of the standardized residuals is shown in the Standardized Regression Residuals section.

```{r}
resid_df <- data.frame(
  predicted = fitted(model_full),
  stdres    = rstandard(model_full)
)

ggplot(resid_df, aes(x = predicted, y = stdres)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, color = "purple", linewidth = 1) +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed") +
  theme_minimal() +
  labs(title = "Standardized Residuals by Predicted Value",
       x = "Predicted LNMEDHVAL", y = "Standardized residual")
```
```{r}
n_out2 <- sum(abs(resid_df$stdres) > 2)
n_out3 <- sum(abs(resid_df$stdres) > 3)
n_out2
n_out3
```
In the plot of standardized residuals against predicted values, 85 block groups (about 4.9%) fall beyond $\pm 2$ and 20 fall beyond $\pm 3$, so there are some outliers. The spread is mostly even but is somewhat wider at the lowest and highest predicted values, which suggests mild heteroscedasticity. Together with the non-linear scatter plots, linearity and homoscedasticity are not fully satisfied.

The maps of the dependent variable and predictors show similar values clustering in space, which suggests spatial autocorrelation and that block groups are not fully independent. The map of the standardized residuals (in the next section) shows whether the residuals are also spatially clustered. We will test spatial autocorrelation formally in the next assignment.

### Additional Models
#### Stepwise Regression
```{r}
step_model <- step(model_full, direction = "both")
step_model$anova
formula(step_model)
```
```{r}
kept_all <- length(coef(step_model)) == length(coef(model_full))
```
Stepwise regression starting from the full model (AIC = -3448.1) did not drop any predictor, because removing any of them would raise the AIC. All four predictors are therefore kept in the final model.

#K-fold Cross Validation

##Sophia
```{r}
#| label: calculating RMSE and re-running the model with two 

cv_data <- regdata %>%
  dplyr::select(LNMEDHVAL, PCTVACANT, PCTSINGLES, PCTBACHMOR, LNBELPOV100, MEDHHINC) %>%
  na.omit()

nrow(cv_data) 

cv_full <- CVlm(
  data    = cv_data,
  form.lm = formula(LNMEDHVAL ~ PCTVACANT + PCTSINGLES + PCTBACHMOR + LNBELPOV100),
  m       = 5,
  seed    = 500,
  plotit  = FALSE,
  printit = FALSE
)

model_two <- lm(LNMEDHVAL ~ PCTVACANT + MEDHHINC, data = cv_data)
summary(model_two)

cv_two <- CVlm(
  data    = cv_data,
  form.lm = formula(LNMEDHVAL ~ PCTVACANT + MEDHHINC),
  m       = 5,
  seed    = 500,
  plotit  = FALSE,
  printit = FALSE
)
rmse <- function(cv) sqrt(mean((cv$LNMEDHVAL - cv$cvpred)^2))
rmse_full <- rmse(cv_full)
rmse_two  <- rmse(cv_two)

rmse_results <- data.frame(
  Model = c("Full model: 4 predictors", "Reduced model: PCTVACANT + MEDHHINC"),
  RMSE  = round(c(rmse_full, rmse_two), 4)
)
knitr::kable(rmse_results)
```
The four-predictor model has a cross-validated RMSE of 0.3682, compared with 0.4435 for the model with only PCTVACANT and MEDHHINC. The four-predictor model has the lower RMSE, so it predicts held-out median house values better. The RMSE is in log-dollar units.

# Standardized Regression Residuals

## Evelyn
```{r}
#| label: Histogram and Chloropleth map of standardized regression residuals
#| message: false
#| warning: false

# save fitted values, residuals, and standardized residuals from the full model
idx <- as.numeric(names(fitted(model_full)))
regdata$predicted <- NA
regdata$resid     <- NA
regdata$stdres    <- NA
regdata$predicted[idx] <- fitted(model_full)
regdata$resid[idx]     <- residuals(model_full)
regdata$stdres[idx]    <- rstandard(model_full)

# histogram of standardized residuals
ggplot(regdata, aes(x = stdres)) +
  geom_histogram(binwidth = 0.25, fill = "#8c6bb1", color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(title = "Histogram of Standardized Residuals",
       x = "Standardized residual", y = "Number of block groups")

# numbers to cite
summary(regdata$stdres)
sum(abs(regdata$stdres) > 2, na.rm = TRUE)  # block groups beyond +/-2
sum(abs(regdata$stdres) > 3, na.rm = TRUE)  # block groups beyond +/-3

# join residuals to the shapefile by POLY_ID
c_map <- c_map %>%
  mutate(POLY_ID = as.numeric(POLY_ID)) %>%
  left_join(regdata %>% dplyr::select(POLY_ID, stdres), by = "POLY_ID")

# choropleth of standardized residuals, same palette as the other maps
ggplot(c_map) +
  geom_sf(aes(fill = stdres)) +
  scale_fill_distiller(palette = "BuPu", direction = 1,
                      limits = c(-3, 3), oob = scales::squish) +
  theme_void() +
  labs(title = "Standardized Regression Residuals", fill = "Std. residual")
```

```

# Discussions and Limitations

## Evelyn
